Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2013-05-22 MIMOSA.tex
Update figure 1 to look nicer

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. proteins, genes, cytokines, etc.) per cell~\citep{Bendall:2011wf}.
Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations. For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point. We simulated data from this constrained model with 200 observations, a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events, with ten independent realizations of data for each $N$. We fit the one-sided MIMOSA model to this data. We evaluated the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C). In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.

>>>>

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. DNA content, cell--surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

We examined the performance of the constrained ($p^{(s)}>p^{(u)}$) and unconstrained ($p^{(s)} \ne p^{(u)}$) beta-binomial mixture models via simulations. For the simulation, we used hyper--parameters estimated from a one-sided MIMOSA model fit to ICS data (IL2 univariate) from the primary immunogenicity time point. We simulated data from this constrained model with 200 observations (subjects), a response rate of 60\%, $N=$ 1,000, 5,000, and 10,000 events (cells), with ten independent realizations of data for each $N$. We fit the one-sided MIMOSA model to this data. We evaluated the sensitivity and specificity of the model's ability to correctly identify observations (subjects) from the ``responder'' and ``non-responder'' groups through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and log fold-change. We repeated this procedure for the two-sided models fit to two-sided data (Figure~\ref{fig:simulations} A,C). In addition, we examined the nominal \textit{vs.} observed FDR to assess the ability of each method to properly estimate the FDR (Figure~\ref{fig:simulations} B,D).

@@

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the event (cell) proportions drawn from  truncated normal distributions with support $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the hyper--parameters  $\alpha$ and $\beta$ estimated from the ICS data set (see Supplementary Figure 3 panels C and D). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test.
